---
title: "CapstoneProject_BellaBeat_GoogleDataAnalytics"
author: "Nico Kosa"
date: "9/5/2021"
output: html_document
---
## EXPANDING VECTOR MEMORY
```{r cars}
Sys.setenv('R_MAX_VSIZE'=32000000000)
```

## LOADING IN PACKAGE LIBRARIES FILLED WITH THE FUNCTIONS AND COMMANDS I WILL USE
```{r cars}
library(tidyverse)
library(readr)
library(lubridate)
library(janitor)
```

## READING THE SPREADHEETS/DATA DOCUMENTS
```{r cars}
secondsHeartrate <- read_csv("Fitabase Data 4.12.16-5.12.16 copy/heartrate_seconds_merged.csv")
hourlyCalories <- read_csv("Fitabase Data 4.12.16-5.12.16 copy/hourlyCalories_merged.csv")
hourlyIntensities <- read_csv("Fitabase Data 4.12.16-5.12.16 copy/hourlyIntensities_merged.csv")
hourlySteps <- read_csv("Fitabase Data 4.12.16-5.12.16 copy/hourlySteps_merged.csv")
minuteMETsNarrow <- read_csv("Fitabase Data 4.12.16-5.12.16 copy/minuteMETsNarrow_merged.csv")
daySleep <- read_csv("Fitabase Data 4.12.16-5.12.16 copy/sleepDay_merged.csv")
```

VIEWING IMPORTED .CSV FILES TO MAKE SURE THEY ARE ACCURATE AND VIEWABLE
```{r cars}
View(secondsHeartrate)
View(hourlyCalories)
View(hourlyIntensities)
View(hourlySteps)
View(minuteMETsNarrow)
View(daySleep)
```

## RENAMING ALL DATE/TIME COLUMNS TO "Time"
```{r cars}
secondsHeartrate <- secondsHeartrate %>% 
  rename(ActivityHour = Time)

minuteMETsNarrow <- minuteMETsNarrow %>% 
  rename(ActivityHour = ActivityMinute)
```

##CHANGING ALL THE DATES AND TIMES TO AN EASIER TO SORT FORMAT AND IN NEW COLUMN TITLED "datetime"
```{r cars}
secondsHeartrate$datetime <- mdy_hms(secondsHeartrate$ActivityHour)
hourlyCalories$datetime <- mdy_hms(hourlyCalories$ActivityHour)
hourlyIntensities$datetime <- mdy_hms(hourlyIntensities$ActivityHour)
hourlySteps$datetime <- mdy_hms(hourlySteps$ActivityHour)
minuteMETsNarrow$datetime <- mdy_hms(minuteMETsNarrow$ActivityHour)
daySleep$datetime <- mdy_hms(daySleep$SleepDay)
```

## DROPPING "ActivityHour" COLUMNS
```{r cars}
secondsHeartrate <- secondsHeartrate[,-2]
hourlyCalories <- hourlyCalories[,-2]
hourlyIntensities <- hourlyIntensities[,-2]
hourlySteps <- hourlySteps[,-2]
minuteMETsNarrow <- minuteMETsNarrow[,-2]
daySleep <- daySleep[,-2]
```

## MERGING
```{r cars}
MainByHour <- merge(hourlyCalories, hourlyIntensities, by = c("Id", "datetime"))
MainByHour <- merge(MainByHour, hourlySteps, by = c("Id", "datetime"))
MainByHour <- merge(MainByHour, minuteMETsNarrow, by = c("Id", "datetime"), all=TRUE)
MainByHour <- merge(MainByHour, secondsHeartrate, by = c("Id", "datetime"), all=TRUE)
MainByHour <- merge(MainByHour, daySleep, by = c("Id", "datetime"), all=TRUE)
```

## EXAMINING MERGE RESULTS
```{r cars}
MainByHour %>% 
  count(Id)

secondsHeartrate %>% 
  count(Id)

minuteMETsNarrow %>% 
  count(Id)

hourlyIntensities %>% 
  count(Id)

hourlySteps %>% 
  count(Id)

hourlyCalories %>% 
  count(Id)

daySleep %>% 
  count(Id)

MainByHour %>% 
  filter(Id=="2022484408")

glimpse(MainByHour)
glimpse(secondsHeartrate)
```
## SUMMARIZING BY HOUR AND GETTING THE HOURLY MEAN FOR NON-TIME VALUES
```{r cars}
MainByHourGrouped <- MainByHour %>% 
  mutate(hour = hour(datetime),
         date = date(datetime)) %>% 
  group_by(Id, date, hour) %>% 
  summarise(Calories = first(Calories),
            TotalIntensity = first(TotalIntensity),
            AverageIntensity = first(AverageIntensity),
            StepTotal = first(StepTotal),
            TotalSleepRecords = first(TotalSleepRecords),
            TotalMinutesAsleep = first(TotalMinutesAsleep),
            TotalTimeInBed = first(TotalTimeInBed),
            METsHour = mean(METs, na.rm=TRUE),
            HeartrateHour = mean(Value, na.rm=TRUE)) 
```

## ANOTHER QUICK GLIMPSE
```{r cars}
MainByHourGrouped %>% 
  filter(Id=="1503960366")
```
## CREATING A NEW DATA FRAME
```{r cars}
hourlyActivity <- MainByHourGrouped %>% 
  arrange(hour) %>% 
  group_by(Id, hour) %>%
  summarise(hourMETs = mean(METsHour, na.rm=TRUE))
```

##DATA VIZ - GETTING EACH PARTICIPANT'S TOTALL METS PER EACH HOUR ACROSS ALL DAYS TO SEE WHICH HOURS HAD THE MOST ACTIVITY
```{r cars}
hourlyActivity %>%
  ggplot(aes(x=hour, y = hourMETs, fill = Id)) +
  geom_area() +
  theme(legend.position="none") +
  scale_x_continuous(name="hour", breaks = c(1, 2, 3, 4, 5, 6, 7, 8 ,9 ,10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23), labels = c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "13", "14", "15", "16", "17", "18", "19", "20", "21", "22", "23"))
```

## SAVING PREVIOUS GGPLOT CHART GRAPHIC
```{r cars}
 ggsave(file="ActivityTimes.png", width=12, height=7)
```

## DATA VIZ - LINE PLOT OF METsHour VS Calories FOR ALL PARTICIPANTS 
```{r cars}
MainByHourGrouped$Id <- as.factor(MainByHourGrouped$Id)

glimpse(MainByHourGrouped)
ggplot(data=MainByHourGrouped) + geom_line(mapping=aes(x=METsHour, y=Calories, color=Id))
```

## SAVING PREVIOUS GGPLOT DATA VIZ
```{r cars}
ggsave(file="metsvscals.png")
```

## GETTING HIGHEST CALORIES/HOUR AND HIGHEST METs/HOUR TO CALCULATE SLOPE
```{r cars}
TopSlope <- MainByHourGrouped %>%
  group_by(Id) %>% 
  mutate(HighestMETsHour = max(METsHour, na.rm=TRUE)) %>%
  distinct(Calories, .keep_all=TRUE) %>%
  mutate(HighestCaloriesHour = max(Calories, na.rm=TRUE)) %>%
  summarize (Id = first(Id), HighestCaloriesHour = first(HighestCaloriesHour), HighestMETsHour = first(HighestMETsHour))
```


## GETTING LOWEST CALORIES/HOUR AND LOWEST METs/HOUR TO CALCULATE SLOPE
```{r cars}
BottomSlope <- MainByHourGrouped %>%
  group_by(Id) %>% 
  filter(METsHour !=0) %>% 
  mutate(LowestMETsHour = min(METsHour, na.rm=TRUE)) %>%
  distinct(Calories, METsHour, .keep_all=TRUE) %>%
  mutate(LowestCaloriesHour = min(Calories, na.rm=TRUE)) %>%
  distinct(METsHour, .keep_all=TRUE) %>%
  mutate(AverageMETs = mean(METsHour, na.rm = TRUE), AverageTimeInBed = mean(TotalTimeInBed, na.rm=TRUE), AverageMinutesAsleep = mean(TotalMinutesAsleep, na.rm=TRUE)) %>% 
  mutate(Id = first(Id), LowestCaloriesHour = first(LowestCaloriesHour), LowestMETsHour = first(LowestMETsHour)) %>% 
  distinct(AverageMETs, .keep_all=TRUE) %>% 
  select(-c(date, hour, Calories, TotalIntensity, AverageIntensity, StepTotal, METsHour, HeartrateHour, TotalSleepRecords, TotalTimeInBed, TotalMinutesAsleep))
```

## COMBINING MIN AND MAX DATA TABLES TO ONE CHART TO CALCULATE SLOPES
```{r cars}
Slopes <- merge(BottomSlope, TopSlope, by = c("Id"))
```

## CALCULATING RISE, RUN, AND SLOPE USING NEW TABLE
```{r cars}
Slopes$Rise <- (Slopes$HighestCaloriesHour - Slopes$LowestCaloriesHour)
Slopes$Run <- (Slopes$HighestMETsHour - Slopes$LowestMETsHour)
Slopes$Slope <- (Slopes$Rise / Slopes$Run)
```

## SORTING IDs/ROWS BY THEIR SLOPE
## TO CREATE A BETTER DATA VIZ
## TO MORE CLEARLY SHOW THE RANGE OF SLOPES
```{r cars}
Slopes %>%
  mutate(Id=fct_reorder(Id, Slope)) %>%
  ggplot(aes(x=Id, y=Slope, fill=Id)) + geom_bar(stat="identity") + theme(legend.position="none", axis.text.x=element_text(angle = -90, hjust = 0))
```

## SAVING PREVIOUS GGPLOT CHART GRAPHIC
```{r cars}
 ggsave(file="Slopes.png", width = 14, height = 7)
```

##ARRANGING THE SLOPES DATAFRAME BY SLOPES
```{r cars}
Slopes %>% 
  arrange(Slope)
```
## GROUPING USERS BASED ON THEIR SLOPES AS INDICATIONS OF THEIR HEALTH LEVELS TO COMPARE USERS
```{r cars}
UsertypeSort <- Slopes %>%
summarise(
user_type = factor(case_when(
    Slope > 7.36 ~ "Less Healthy",
    Slope < 7.36 & Slope > 5.64 ~ "Medium Healthy",
    Slope < 5.64 ~ "More Healthy"
),levels=c("Less Healthy", "Medium Healthy", "More Healthy")), AverageMETs, AverageTimeInBed, AverageMinutesAsleep,Slope,  Id=Id)
```


##MAKING SURE EACH USERTYPE HAS AN EQUAL NUMBER OF PARTICIPANTS
```{r cars}
UsertypeSort %>% 
  arrange(Slope)
```

## ADDING AN AVERAGE TIME AWAKE IN BED COLUMN
```{r cars}
UsertypeSort$AvgTimeAwakeInBed <- (UsertypeSort$AverageTimeInBed - UsertypeSort$AverageMinutesAsleep)
```

## DATA VIZ - COMPARING AVERAGE METs OF USER GROUPS
```{r cars}
ggplot(UsertypeSort, aes(user_type, AverageMETs, fill=user_type)) +
  geom_boxplot() +
  theme(legend.position="none")
```

## SAVING DATA VIZ
```{r cars}
ggsave(file = "UserTypeMETs.png")
```

## DATA VIZ - COMPARING AVERAGE TIME ASLEEP OF USER GROUPS
```{r cars}
ggplot(UsertypeSort, aes(user_type, AverageMinutesAsleep, fill=user_type)) +
  geom_boxplot() +
  theme(legend.position="none")
```
## SAVING DATA VIZ
```{r cars}
ggsave(file="UserTypeSleepTime.png")
```
## DATA VIZ - COMPARING AVERAGE TIME AWAKE IN BED OF USER GROUPS
```{r cars}
ggplot(UsertypeSort, aes(user_type, AvgTimeAwakeInBed, fill=user_type)) +
  geom_boxplot() +
  theme(legend.position="none")
```

## SAVING DATA VIZ
```{r cars}
ggsave(file="UserTypeTimeInBedAwake.png")
```